---
title: Clean and organize trait data
author: Arun Chavan
format: 
  html:
    theme: cosmo
    self-contained: true
toc: true
editor: source
fontsize: 0.95em
---

Started: 2023-11-20  
Last edited: `r format(Sys.time())`

# Setup
```{r}
#| message: false
#| warning: false

knitr::opts_chunk$set(dpi = 300, collapse = TRUE, comment = "#>", error = TRUE)

library(tidyverse)
library(here)
library(readxl)
```

# Data
```{r}
#| warning: false
#| message: false

# traits
kd <- read_excel(here("data/traits/MP-KD/MammaliaData_MP_KD_20161207.xlsx"), 
                 na = c("", "NA"))
pt <- read_tsv(here("data/traits/PanTHERIA/PanTHERIA_1-0_WR05_Aug2008.txt"),
               na = c(-999, "-999", "", "NA"), guess_max = 10000)
mv <- read_csv(here("data/traits/Myhrvold/Amniote_Database_Aug_2015.csv"), 
               na = "-999", guess_max = 10000)

# list of mammalian orders 
source(here("code/utilities_mammalian-orders.R"))
```

Keep relevant variables and clean up: 

```{r}
pt <- pt %>% 
  select(order = MSW05_Order, 
         family = MSW05_Family, 
         genus = MSW05_Genus, 
         species = MSW05_Species, 
         binomial = MSW05_Binomial,
         adult_body_mass_g = `5-1_AdultBodyMass_g`,
         gestation_len_d = `9-1_GestationLen_d`,
         litter_size = `15-1_LitterSize`,
         age_at_eye_opening_d = `2-1_AgeatEyeOpening_d`,
         weaning_age_d = `25-1_WeaningAge_d`,
         neonate_body_mass_g = `5-3_NeonateBodyMass_g`)

pt$binomial <- str_replace_all(pt$binomial, "\\s+", "_")

# replacing -999 with NA did not work in read_tsv
replace_999 <- function(x) if_else(x == -999, NA, x)

pt <- pt %>% 
  mutate(across(where(is.double), replace_999))
```

```{r}
mv <- mv %>% 
  filter(class == "Mammalia") %>% 
  select(order,
         family,
         genus, 
         species,
         common_name, 
         litter_size = litter_or_clutch_size_n,
         adult_body_mass_g,
         gestation_len_d = gestation_d,
         weaning_age_d = weaning_d,
         neonate_body_mass_g = birth_or_hatching_weight_g) %>% 
  mutate(binomial = paste0(genus, "_", species)) %>% 
  relocate(binomial, .after = species)
```

```{r}
kd <- kd %>% 
  select(binomial = Species,
         adult_body_mass_g = AdultBodyMass_g,
         gestation_len_d = GestationLen_d,
         age_at_eye_opening_d = AgeatEyeOpening_d,
         litter_size = LitterSize,
         precocity_kd = PrecocityAtBirth) %>% 
  mutate(precocity_kd = str_to_lower(precocity_kd))
```

# Consolidate taxonomic information

First create a list of all binomials with their taxonomic information from both PanTHERIA and Myhrvold. 

```{r}
spp <- bind_rows(myhrvold = select(mv, order, family, binomial),
                 pantheria = select(pt, order, family, binomial), 
                 .id = "source") %>% 
  arrange(binomial)
```

## Resolve discordances between the two datasets

If the family and order assignment between the datasets is the same for a given species, merge the information; if not, resolve the discordance. 

```{r}
spp_wide <- spp %>% 
  pivot_wider(id_cols = "binomial", 
              names_from = "source", 
              values_from = c("order", "family"))
```

For species that are present in only dataset, we will use the order and family information from whichever dataset they are present in. 

For species that are present in both datasets, but the order or family information is discordant between the two datasets, we need to resolve the discordance. There are only 10 such species, below. 

```{r}
spp_wide %>% 
  filter(!is.na(order_myhrvold) & !is.na(order_pantheria)) %>% # present in both
  filter(order_myhrvold != order_pantheria | 
           family_myhrvold != family_pantheria)
```

The discordance is actually due to a misspelling of Indriidae ("The Indriidae (sometimes incorrectly spelled Indridae) are a family of strepsirrhine primates." from https://en.wikipedia.org/wiki/Indriidae). We just need to correct this spelling to resolve all discordances. 

```{r}
spp_wide <- spp_wide %>% 
  mutate(family_myhrvold = str_replace(family_myhrvold, "Indridae", "Indriidae"),
         family_pantheria = str_replace(family_pantheria, "Indridae", "Indriidae"))
```

Let's make sure that now we have no discordance. 

```{r}
spp_wide %>% 
  filter(!is.na(order_myhrvold) & !is.na(order_pantheria)) %>% # present in both
  filter(order_myhrvold != order_pantheria | 
           family_myhrvold != family_pantheria)
```

## Combine taxonomic information from both datasets

Now that we have resolved the discordance, we will keep only one column for order and one for family by coalescing information from the two datasets. 

```{r}
spp_resolved <- spp_wide %>% 
  mutate(order = coalesce(order_pantheria, order_myhrvold),
         family = coalesce(family_pantheria, family_myhrvold)) %>% 
  select(binomial, order, family)
```

## Add infraclass and superorder information

```{r}
# turn the order list into a data frame
mammal_orders <- bind_rows(
  (enframe(Mammalia[[c("Eutheria")]],
           name = "superorder", 
           value = "order") %>% 
     unnest_longer(order) %>% 
     mutate(infraclass = "Eutheria")),
  (enframe(Mammalia[c("Prototheria", "Metatheria")], 
           name = "infraclass", 
           value = "order") %>% 
     unnest_longer(order))
) %>% 
  relocate(infraclass)
```

```{r}
spp_resolved <- left_join(spp_resolved, mammal_orders, by = "order") %>% 
  select(infraclass, superorder, order, family, binomial) %>% 
  arrange(infraclass, superorder, order, family, binomial)
```

# Add taxonomic information to the trait datasets

We will now add the consolidated taxonomic information to all three trait datasets. 
```{r}
pt <- pt %>% 
  select(-order, -family, -genus, -species) %>% 
  left_join(spp_resolved, by = "binomial") %>% 
  relocate(infraclass, superorder, order, family)

mv <- mv %>% 
  select(-order, -family, -genus, -species) %>% 
  left_join(spp_resolved, by = "binomial") %>% 
  relocate(infraclass, superorder, order, family)

kd <- kd %>% 
  left_join(spp_resolved, by = "binomial") %>% 
  relocate(infraclass, superorder, order, family)
```

# Write files

```{r}
# trait data
fs::dir_create(here("data/01_cleaned"))
write_csv(pt, here("data/01_cleaned/pantheria.csv"))
write_csv(mv, here("data/01_cleaned/myhrvold.csv"))
write_csv(kd, here("data/01_cleaned/mp-kd.csv"))

# taxonomic information
write_csv(spp_resolved, here("data/taxonomic-information.csv"))
```

# Session Info
```{r session_info, eval=TRUE}
sessionInfo()
```
